The Rockerverse: Packages and Applications for Containerisation with R

### 2020-03-02 manuscript.Rmd
closes #71 (together with previously merged PR)
update CRAN references, closes #63

**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor)) is no longer actively maintained, but it should be honorably mentioned as the first R package for managing Docker images and containers.
It uses the \CRANpkg{sys} package to run system commands against the Docker CLI, both locally and through an SSH connection, and it has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
The outputs of container executions are converted to appropriate R types.
The Docker CLI's basic functionality, although it evolves quickly and with little concern for avoiding breaking changes, has remained unchanged in core functions, meaning that a core function such as `harbor::docker_run(image = "hello-world")` still works depsite its stopped development.
```{r harbor, eval=FALSE, include=FALSE}
remotes::install_github("wch/harbor")
harbor::docker_run(image = "hello-world")
```

**\CRANpkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) is currently the most powerful Docker client in R.
It interfaces with the Docker daemon over the Docker HTTP API\footnote{\href{https://docs.docker.com/engine/api/latest/}{https://docs.docker.com/engine/api/latest/}} via a Unix socket on Linux or MacOS, over a named pipe on Windows, or over an HTTP/TCP connection.
The package is the only one not using system calls to the `docker` CLI tool for managing images and containers.
The package thereby enables connections to remote Docker instances without direct configuration of the local Docker daemon.
Furthermore using the API gives access to information in a structured way, is system independent, and is likely more reliable than parsing command line output.
\CRANpkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but it is still similar to the Docker CLI.
The interface is similar to R6 objects, in that a `stevedore_object` has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.

@@

The _Bioconductor_ `Dockerfile` selects the desired version of R from Rocker, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages.
A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
However, a recent audit of the _Bioconductor_  image stack `Dockerfile` led to the deprecation of several community-maintained images, because the numerous specific images became too hard to understand, complex to maintain, and cumbersome to customise.
As part of the simplification, a recent innovation is to produce a `bioconductor_docker:devel` image to emulate the _Bioconductor_ nightly Linux build machine as closely as possible.
This image contains the build system environment variables and the _system dependencies_ needed to install and check almost all (1813 out of 1823) _Bioconductor_ software packages, and it saves users and package developers from managing these themselves.
Furthermore, the image is configured so that `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location.
Users mounting a location on the host file system to this location can persistently manage installed packages across Docker containers or image updates.
Many R users pursue flexible work flows tailored to particular analysis needs rather than standardized work flows.
The new `bioconductor_docker` image is well suited for this preference, while `bioconductor_docker:devel` provides developers with a test environment close to _Bioconductor_'s build system.

@@

It includes machine learning libraries such as Tensorflow and Keras (see also image `rocker/ml` in Section&nbsp;\nameref{rocker-gpu}), and it also configures the \CRANpkg{reticulate} package.

@@

- The [_Radiant_](https://radiant-rstats.github.io/docs/) project provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \CRANpkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.

@@

**\pkg{dockerfiler}** ([https://github.com/ColinFay/dockerfiler/](https://github.com/ColinFay/dockerfiler/)) is an R package designed for building `Dockerfile`s straight from R.
A scripted creation of a `Dockerfile` enables iteration and automation, for example for packaging applications for deployment (see \nameref{deployment}).
Developers can retrieve system requirements and package dependencies to write a `Dockerfile`, for example, by leveraging the tools available in R to parse a `DESCRIPTION` file.

@@

**\pkg{dockr}** ([https://github.com/smaakage85/dockr](https://github.com/smaakage85/dockr)) is a similar package focusing on the generation of Docker images for R packages, in which the package itself and all of the R dependencies, including local non-CRAN packages, are available \citep{kjeldgaard_dockr_2019}.
\pkg{dockr} facilitates the organisation of code in the R package structure and the resulting Docker image mirrors the package versions of the current R session.
Users can manually add statements for non-R dependencies to the `Dockerfile`.

@@

System dependencies are the domain of Docker, but for a full description of the computing environment one must also manage the R version and the the R packages.
R versions are available via the versioned Rocker image stack \citep{RJ-2017-065}.
[r-online](https://github.com/ColinFay/ronline) leverages these images and provides an app for helping users to detect breaking changes between different R versions and for historic exploration of R.
With a standalone NodeJS app or [r-online](https://srv.colinfay.me/r-online), the user can compare a piece of code run in two separate versions of R.
Internally, r-online opens one or two Docker instances with the given version of R based on Rocker images, executes a given piece of code, and returns the result to the user.
Regarding R package management, this can be achieved with MRAN, or with packages such as \CRANpkg{checkpoint} and **\CRANpkg{renv}**, which can naturally be applied within images and containers.
For example, \CRANpkg{renv} ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) helps users to manage the state of the R library in a reproducible way, further providing isolation and portability \citep{renv2019}.
While \CRANpkg{renv} does not cover system dependencies, the \CRANpkg{renv}-based environment can be transferred into a container either by restoring the environment based on the main configuration file `renv.lock` or by storing the \CRANpkg{renv}-cache on the host and not in the container \citep{ushey_using_2019}.
With both the system dependencies and R packages consciously managed in a Docker image, users can start using containers as the _only_ environment for their workflows, which allows them to work indepenently of physical computers\footnote{Allowing them to be digital "nomads", cf. J.~Bryan's \href{https://github.com/jennybc/docker-why}{https://github.com/jennybc/docker-why}.} and to assert a specific degree of confidence in the stability of a developed software \citep[cf. \texttt{README.Rmd} in][]{marwick_research_2017}.

@@

Second, the strong integration of **system libraries in core packages** in the [R-spatial community](https://www.r-spatial.org/) makes containers essential for stable and proactive development of common classes for geospatial data modelling and analysis. For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats.
\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and, therefore, the maintainers spend a lot of effort trying not to break them.
Purpose-built Docker images are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf. \href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
```{r geospatial_deps, eval=FALSE, include=FALSE}
length(tools::dependsOnPkgs(c("rgdal", "sf"), dependencies = c("most"), recursive = TRUE))
```

@@

\label{ci}
Fifth, Docker images can be used **on continuous integration (CI) platforms** to streamline the testing of packages.
\citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
\citet{cardozo_faster_2018} also saved time with Travis&nbsp;CI by re-using the testing image as the basis for an image intended for publication on Docker&nbsp;Hub.
[`r-ci`](https://github.com/ColinFay/r-ci) is, in turn, used with [GitLab&nbsp;CI](https://docs.gitlab.com/ee/ci/), which itself is built on top of Docker images: the user specifies a base Docker image and control code, and the whole set of tests is run inside a container.
The `r-ci` image stack combines `rocker` versioning and a series of tools specifically designed for testing in a fixed environment with a customized list of preinstalled packages.
Especially for long-running tests or complex system dependencies, these approaches to seperate installation of build dependencies with code testing streamline the development process.
Furthermore, with each change, even this manuscript is rendered into a PDF and deployed to a GitHub-hosted website (see `.travis.yml` and `Dockerfile` in the manuscript repository), not because of concern about time, but to control the environment used on a CI server.
This gives, on the one hand, easy access after every update of the R Markdown source code and, on the other hand a second controlled environment to make sure that the article renders successfully and correctly.

@@

\CRANpkg{googleComputeEngineR} integrates with the R parallelisation package \CRANpkg{future} to enable this with only a few lines or R code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.

@@

This can be used to scale up R code to millions of instances if need be with little or no changes to existing code, as demonstrated by the proof of concept `cloudRunR`\footnote{\href{https://github.com/MarkEdmondson1234/cloudRunR}{https://github.com/MarkEdmondson1234/cloudRunR}}, which uses Cloud Run to create a scalable R-based API using \CRANpkg{plumber}.

@@

First, different packages use containers for the **deployment of R and \CRANpkg{Shiny} apps**.
\CRANpkg{Shiny} is a popular package for creating interactive online dashboards with R, and it enables users with very diverse backgrounds to create stable and user-friendly web applications.
_ShinyProxy_ ([https://www.shinyproxy.io/](https://www.shinyproxy.io/)) is an open source tool to deploy Shiny apps in an enterprise context, where it features single sign-on, but it can also be used in scientific use cases \citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
ShinyProxy uses Docker containers to isolate user sessions and to achieve scalability for multi-user scenarious with multiple apps.
ShinyProxy itself is written in Java to accomodate corporate requirements and may itself run in a container for stability and availability.
The tool is built on ContainerProxy ([https://www.containerproxy.io/](https://www.containerproxy.io/)), which provides similar features for executing long-running R jobs or interactive R sessions.
The started containers can run on a regular Docker host but also in clusters.
Continuous integration and deployment (CI/CD) for Shiny applications using Shinyproxy can be achieved, e.g., via [GitLab pipelines](https://docs.gitlab.com/ee/ci/pipelines.html).
A pipeline can include building and checking R packages and Shiny apps.
After the code has passed the checks, Docker images are built and pushed to GitLab's container registry.
The pipeline finishes with triggering a webhook on the server, where the
deployment script is executed.
The script updates configurations, pulls the new Docker images, and restarts the services.

Another example is the package \CRANpkg{golem} ([https://github.com/ThinkR-open/golem](https://github.com/ThinkR-open/golem)), which makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building and deploying production-grade Shiny applications.
\CRANpkg{googleComputeEngineR} ([https://cloudyr.github.io/googleComputeEngineR/](https://cloudyr.github.io/googleComputeEngineR/)) enables quick deployments of key R services, such as RStudio and Shiny, onto cloud virtual machines (VMs) with Google Cloud Compute Engine \citep{googleComputeEngineR_2019}.
The package utilises `Dockerfile`s to move the labour of setting up those services from the user to a premade Docker image, which is configured and run in the cloud VM.
For example, by specifying the template `template="rstudio"` in functions `gce_vm_template()/gce_vm()` an up-to-date RStudio Server image is launched for development work, whereas specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.

@@

R has been historically viewed as a tool for analysis and scientific research, but not for creating software that corporations can rely on for production services.
However, thanks to advancements in R running as a web service, along with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
A number of packages and projects have enabled R to respond to client reqests over TCP/IP and local socket servers, such as \CRANpkg{Rserve}, \CRANpkg{svSocket}, [rApache](http://www.rapache.net) and more recently \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) and \pkg{RestRserve} ([http://restrserve.org](http://restrserve.org)), which even processes incoming requests in parallel with forked processes using \CRANpkg{Rserve}.
The latter two also provide documentation for deployment with Docker or ready-to-use images with automated builds\footnote{See \href{https://www.rplumber.io/docs/hosting.html\#docker}{https://www.rplumber.io/docs/hosting.html\#docker}, \href{https://hub.docker.com/r/trestletech/plumber/}{https://hub.docker.com/r/trestletech/plumber/} and \href{https://hub.docker.com/r/rexyai/restrserve/}{https://hub.docker.com/r/rexyai/restrserve/}.}.
These software allow other (remote) processes and programming languages to interact with R and to expose R-based function in a service architecture with HTTP APIs.
APIs based on these packages can be deployed with scalability and high availability using containers.
This pattern of deploying code matches those used by software engineering services created in more established languages in the enterprise domain, such as Java or Python, and R can be used alongside those langages as a first-class member of a software engineering technical stack.

@@

The AI @ T-Mobile team created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers \citep{t-mobile_enterprise_2018}.
For example, one model quickly identifies whether a message is from a customer \citep[see \CRANpkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo} further described by ][]{nolis_small_2019}, and others tell which customers are likely to make a repeat purchase.
If a data scientist creates a machine learning model and exposes it through a \CRANpkg{plumber} API, then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
The models are convolutional neural networks that use the \CRANpkg{keras} package and run in a Rocker Docker container.
The corresponding `Dockerfile`s are published [on GitHub](https://github.com/tmobile/r-tensorflow-api).
Since the models power tools for agents and customers, they need to have extremely high uptime and reliability.
The AI @ T-Mobile team found that the models performed well, and today these models power real-time services that are called over a million times a day.

@@

**Teaching** is a further example of where shared computing environments and sandboxing can greatly improve the process.
First, _prepared environments for teaching_ are especially helpful for (a) introductory courses, where students often struggle with the first step of installation and configuration \citep{cetinkaya-rundel_infrastructure_2018}, and (b) courses that require access to a relatively complex setup of software tools, e.g., database systems.
\citet{cetinkaya-rundel_infrastructure_2018} describe how a Docker-based deployment of RStudio (i) avoided problems with troubleshooting individual students' computers and greatly increased engagement through very quickly showing tangible outcomes, e.g., a visualisation, and (ii) reduced demand on teaching and IT staff.
Each student received access to a personal RStudio instance running in a container after authentication with the university login, which gives the benefits of sandboxing and the possibility of limiting resources.
\citet{cetinkaya-rundel_infrastructure_2018} found that for the courses at hand, actual usage of the UI is intermittent so a single cloud-based VM with four cores and 28&nbsp;GB RAM sufficed for over 100 containers.
An example for mitigating _complex setups_ is teaching databases.
R is very useful tool for interfacing with databases, because almost every open source and proprietary database system has an R package that allows users to connect and interact with it.
This flexibility is even broadened by \CRANpkg{DBI}, which allows for creating a common API for interfacing these databases, or the \CRANpkg{dbplyr} package, which runs \CRANpkg{dplyr} code straight against the database as queries.
But learning and teaching these tools comes with the cost of deploying or having access to an environment with the software and drivers installed.
For people teaching R, it can become a barrier if they need to install local versions of database drivers or connect to remote instances which might or might not be made available by IT services.
Giving access to a sandbox for the most common environments for teaching databases is the idea behind [`r-db`](https://github.com/ColinFay/r-db), a Docker image that contains everything needed to connect to a database from R.
Notably, with `r-db`, users do not have to install complex drivers or configure their machine in a specific way.
The `rocker/tidyverse` base image ensures that users can also readily use packages for analysis, display, and reporting.

>>>>

**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor)) is no longer actively maintained, but it should be honorably mentioned as the first R package for managing Docker images and containers.
It uses the \CRANpkg{sys} package \citep{cran_sys} to run system commands against the Docker CLI, both locally and through an SSH connection, and it has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
The outputs of container executions are converted to appropriate R types.
The Docker CLI's basic functionality, although it evolves quickly and with little concern for avoiding breaking changes, has remained unchanged in core functions, meaning that a core function such as `harbor::docker_run(image = "hello-world")` still works depsite its stopped development.
```{r harbor, eval=FALSE, include=FALSE}
remotes::install_github("wch/harbor")
harbor::docker_run(image = "hello-world")
```

**\CRANpkg{stevedore}** is currently the most powerful Docker client in R \citep{cran_stevedore}.
It interfaces with the Docker daemon over the Docker HTTP API\footnote{\href{https://docs.docker.com/engine/api/latest/}{https://docs.docker.com/engine/api/latest/}} via a Unix socket on Linux or MacOS, over a named pipe on Windows, or over an HTTP/TCP connection.
The package is the only one not using system calls to the `docker` CLI tool for managing images and containers.
The package thereby enables connections to remote Docker instances without direct configuration of the local Docker daemon.
Furthermore using the API gives access to information in a structured way, is system independent, and is likely more reliable than parsing command line output.
\CRANpkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but it is still similar to the Docker CLI.
The interface is similar to R6 objects, in that a `stevedore_object` has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.

@@

The _Bioconductor_ `Dockerfile` selects the desired version of R from Rocker, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages \citep{cran_biocmanager}.
A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
However, a recent audit of the _Bioconductor_  image stack `Dockerfile` led to the deprecation of several community-maintained images, because the numerous specific images became too hard to understand, complex to maintain, and cumbersome to customise.
As part of the simplification, a recent innovation is to produce a `bioconductor_docker:devel` image to emulate the _Bioconductor_ nightly Linux build machine as closely as possible.
This image contains the build system environment variables and the _system dependencies_ needed to install and check almost all (1813 out of 1823) _Bioconductor_ software packages, and it saves users and package developers from managing these themselves.
Furthermore, the image is configured so that `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location.
Users mounting a location on the host file system to this location can persistently manage installed packages across Docker containers or image updates.
Many R users pursue flexible work flows tailored to particular analysis needs rather than standardized work flows.
The new `bioconductor_docker` image is well suited for this preference, while `bioconductor_docker:devel` provides developers with a test environment close to _Bioconductor_'s build system.

@@

It includes machine learning libraries such as Tensorflow and Keras (see also image `rocker/ml` in Section&nbsp;\nameref{rocker-gpu}), and it also configures the \CRANpkg{reticulate} package \citep{cran_reticulate}.

@@

- The [_Radiant_](https://radiant-rstats.github.io/docs/) project provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \CRANpkg{Shiny} \citep{cran_shiny}, and for use in education as part of an MSc course\footnote{`Dockerfile` available on GitHub: \href{https://github.com/radiant-rstats/docker}{https://github.com/radiant-rstats/docker}.}.

@@

**\CRANpkg{dockerfiler}** is an R package designed for building `Dockerfile`s straight from R \citep{cran_dockerfiler}.
A scripted creation of a `Dockerfile` enables iteration and automation, for example for packaging applications for deployment (see \nameref{deployment}).
Developers can retrieve system requirements and package dependencies to write a `Dockerfile`, for example, by leveraging the tools available in R to parse a `DESCRIPTION` file.

@@

**\CRANpkg{dockr}** is a similar package focusing on the generation of Docker images for R packages, in which the package itself and all of the R dependencies, including local non-CRAN packages, are available \citep{cran_dockr,kjeldgaard_dockr_2019}.
\pkg{dockr} facilitates the organisation of code in the R package structure and the resulting Docker image mirrors the package versions of the current R session.
Users can manually add statements for non-R dependencies to the `Dockerfile`.

@@

System dependencies are the domain of Docker, but for a full description of the computing environment one must also manage the R version and the the R packages.
R versions are available via the versioned Rocker image stack \citep{RJ-2017-065}.
[r-online](https://github.com/ColinFay/ronline) leverages these images and provides an app for helping users to detect breaking changes between different R versions and for historic exploration of R.
With a standalone NodeJS app or [r-online](https://srv.colinfay.me/r-online), the user can compare a piece of code run in two separate versions of R.
Internally, r-online opens one or two Docker instances with the given version of R based on Rocker images, executes a given piece of code, and returns the result to the user.
Regarding R package management, this can be achieved with MRAN, or with packages such as \CRANpkg{checkpoint} \citep{cran_checkpoint} and \CRANpkg{renv} \citep{renv2019}, which can naturally be applied within images and containers.
For example, \CRANpkg{renv} helps users to manage the state of the R library in a reproducible way, further providing isolation and portability.
While \CRANpkg{renv} does not cover system dependencies, the \CRANpkg{renv}-based environment can be transferred into a container either by restoring the environment based on the main configuration file `renv.lock` or by storing the \CRANpkg{renv}-cache on the host and not in the container \citep{ushey_using_2019}.
With both the system dependencies and R packages consciously managed in a Docker image, users can start using containers as the _only_ environment for their workflows, which allows them to work indepenently of physical computers\footnote{Allowing them to be digital "nomads", cf. J.~Bryan's \href{https://github.com/jennybc/docker-why}{https://github.com/jennybc/docker-why}.} and to assert a specific degree of confidence in the stability of a developed software \citep[cf. \texttt{README.Rmd} in][]{marwick_research_2017}.

@@

Second, the strong integration of **system libraries in core packages** in the [R-spatial community](https://www.r-spatial.org/) makes containers essential for stable and proactive development of common classes for geospatial data modelling and analysis. For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats \citep{pebesma_simple_2018,cran_rgdal}.
\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and, therefore, the maintainers spend a lot of effort trying not to break them.
Purpose-built Docker images are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf. \href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
```{r geospatial_deps, eval=FALSE, include=FALSE}
length(tools::dependsOnPkgs(c("rgdal", "sf"), dependencies = c("most"), recursive = TRUE))
```

@@

\label{ci}
Fifth, Docker images can be used **on continuous integration (CI) platforms** to streamline the testing of packages.
\citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
\citet{cardozo_faster_2018} also saved time with Travis&nbsp;CI by re-using the testing image as the basis for an image intended for publication on Docker&nbsp;Hub.
[`r-ci`](https://github.com/ColinFay/r-ci) is, in turn, used with [GitLab&nbsp;CI](https://docs.gitlab.com/ee/ci/), which itself is built on top of Docker images: the user specifies a base Docker image and control code, and the whole set of tests is run inside a container.
The `r-ci` image stack combines `rocker` versioning and a series of tools specifically designed for testing in a fixed environment with a customized list of preinstalled packages.
Especially for long-running tests or complex system dependencies, these approaches to seperate installation of build dependencies with code testing streamline the development process.
Containers can also simplify the integration of R software into larger, multi-language CI pipelines.
Furthermore, with each change, even this manuscript is rendered into a PDF and deployed to a GitHub-hosted website (see `.travis.yml` and `Dockerfile` in the manuscript repository), not because of concern about time, but to control the environment used on a CI server.
This gives, on the one hand, easy access after every update of the R Markdown source code and, on the other hand a second controlled environment to make sure that the article renders successfully and correctly.

@@

\CRANpkg{googleComputeEngineR} integrates with the R parallelisation package \CRANpkg{future} \citep{cran_future} to enable this with only a few lines or R code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.

@@

This can be used to scale up R code to millions of instances if need be with little or no changes to existing code, as demonstrated by the proof of concept `cloudRunR`\footnote{\href{https://github.com/MarkEdmondson1234/cloudRunR}{https://github.com/MarkEdmondson1234/cloudRunR}}, which uses Cloud Run to create a scalable R-based API using \CRANpkg{plumber} \citep{cran_plumber}.

@@

First, different packages use containers for the **deployment of R and \CRANpkg{Shiny} apps**.
\CRANpkg{Shiny} is a popular package for creating interactive online dashboards with R, and it enables users with very diverse backgrounds to create stable and user-friendly web applications \citep{cran_shiny}.
_ShinyProxy_ ([https://www.shinyproxy.io/](https://www.shinyproxy.io/)) is an open source tool to deploy Shiny apps in an enterprise context, where it features single sign-on, but it can also be used in scientific use cases \citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
ShinyProxy uses Docker containers to isolate user sessions and to achieve scalability for multi-user scenarious with multiple apps.
ShinyProxy itself is written in Java to accomodate corporate requirements and may itself run in a container for stability and availability.
The tool is built on ContainerProxy ([https://www.containerproxy.io/](https://www.containerproxy.io/)), which provides similar features for executing long-running R jobs or interactive R sessions.
The started containers can run on a regular Docker host but also in clusters.
Continuous integration and deployment (CI/CD) for Shiny applications using Shinyproxy can be achieved, e.g., via [GitLab pipelines](https://docs.gitlab.com/ee/ci/pipelines.html).
A pipeline can include building and checking R packages and Shiny apps.
After the code has passed the checks, Docker images are built and pushed to GitLab's container registry.
The pipeline finishes with triggering a webhook on the server, where the
deployment script is executed.
The script updates configurations, pulls the new Docker images, and restarts the services.

Another example is the package \CRANpkg{golem}, which makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building and deploying production-grade Shiny applications \citep{cran_golem}.
\CRANpkg{googleComputeEngineR} enables quick deployments of key R services, such as RStudio and Shiny, onto cloud virtual machines (VMs) with Google Cloud Compute Engine \citep{googleComputeEngineR_2019}.
The package utilises `Dockerfile`s to move the labour of setting up those services from the user to a premade Docker image, which is configured and run in the cloud VM.
For example, by specifying the template `template="rstudio"` in functions `gce_vm_template()/gce_vm()` an up-to-date RStudio Server image is launched for development work, whereas specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.

@@

R has been historically viewed as a tool for analysis and scientific research, but not for creating software that corporations can rely on for production services.
However, thanks to advancements in R running as a web service, along with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
A number of packages and projects have enabled R to respond to client reqests over TCP/IP and local socket servers, such as \CRANpkg{Rserve} \citep{cran_rserve}, \CRANpkg{svSocket} \citep{grosjean_sciviews_2019}, [rApache](http://www.rapache.net) and more recently \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) and \pkg{RestRserve} ([http://restrserve.org](http://restrserve.org)), which even processes incoming requests in parallel with forked processes using \CRANpkg{Rserve}.
The latter two also provide documentation for deployment with Docker or ready-to-use images with automated builds\footnote{See \href{https://www.rplumber.io/docs/hosting.html\#docker}{https://www.rplumber.io/docs/hosting.html\#docker}, \href{https://hub.docker.com/r/trestletech/plumber/}{https://hub.docker.com/r/trestletech/plumber/} and \href{https://hub.docker.com/r/rexyai/restrserve/}{https://hub.docker.com/r/rexyai/restrserve/}.}.
These software allow other (remote) processes and programming languages to interact with R and to expose R-based function in a service architecture with HTTP APIs.
APIs based on these packages can be deployed with scalability and high availability using containers.
This pattern of deploying code matches those used by software engineering services created in more established languages in the enterprise domain, such as Java or Python, and R can be used alongside those langages as a first-class member of a software engineering technical stack.

@@

The AI @ T-Mobile team created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers \citep{t-mobile_enterprise_2018}.
For example, one model quickly identifies whether a message is from a customer \citep[see \CRANpkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo} further described by ][]{nolis_small_2019}, and others tell which customers are likely to make a repeat purchase.
If a data scientist creates a machine learning model and exposes it through a \CRANpkg{plumber} API, then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
The models are convolutional neural networks that use the \CRANpkg{keras} package \citep{cran_keras} and run in a Rocker Docker container.
The corresponding `Dockerfile`s are published [on GitHub](https://github.com/tmobile/r-tensorflow-api).
Since the models power tools for agents and customers, they need to have extremely high uptime and reliability.
The AI @ T-Mobile team found that the models performed well, and today these models power real-time services that are called over a million times a day.

@@

**Teaching** is a further example of where shared computing environments and sandboxing can greatly improve the process.
First, _prepared environments for teaching_ are especially helpful for (a) introductory courses, where students often struggle with the first step of installation and configuration \citep{cetinkaya-rundel_infrastructure_2018}, and (b) courses that require access to a relatively complex setup of software tools, e.g., database systems.
\citet{cetinkaya-rundel_infrastructure_2018} describe how a Docker-based deployment of RStudio (i) avoided problems with troubleshooting individual students' computers and greatly increased engagement through very quickly showing tangible outcomes, e.g., a visualisation, and (ii) reduced demand on teaching and IT staff.
Each student received access to a personal RStudio instance running in a container after authentication with the university login, which gives the benefits of sandboxing and the possibility of limiting resources.
\citet{cetinkaya-rundel_infrastructure_2018} found that for the courses at hand, actual usage of the UI is intermittent so a single cloud-based VM with four cores and 28&nbsp;GB RAM sufficed for over 100 containers.
An example for mitigating _complex setups_ is teaching databases.
R is very useful tool for interfacing with databases, because almost every open source and proprietary database system has an R package that allows users to connect and interact with it.
This flexibility is even broadened by \CRANpkg{DBI} \citep{cran_dbi}, which allows for creating a common API for interfacing these databases, or the \CRANpkg{dbplyr} package \citep{cran_dbplyr}, which runs \CRANpkg{dplyr} \citep{cran_dplyr} code straight against the database as queries.
But learning and teaching these tools comes with the cost of deploying or having access to an environment with the software and drivers installed.
For people teaching R, it can become a barrier if they need to install local versions of database drivers or connect to remote instances which might or might not be made available by IT services.
Giving access to a sandbox for the most common environments for teaching databases is the idea behind [`r-db`](https://github.com/ColinFay/r-db), a Docker image that contains everything needed to connect to a database from R.
Notably, with `r-db`, users do not have to install complex drivers or configure their machine in a specific way.
The `rocker/tidyverse` base image ensures that users can also readily use packages for analysis, display, and reporting.

Second, the idea of a common environment and partitioning allows for using _containers in teaching for secure execution and automated testing_ of submissions by students.
[Dodona](https://dodona.ugent.be) is a web platform developed at Ghent University that is used to teach students basic programming skills, and it uses Docker containers to test submissions by students.
This means that both the code testing the students' submissions and the submission itself are executed in a predictable environment, avoiding compatibility issues between the wide variety of configurations used by students.
The containerisation is also used to shield the Dodona servers from bad or even malicious code: memory, time and I/O&nbsp;limits are used to make sure students cannot overload the system.
The web application managing the containers communicates with them by sending configuration information as a JSON document over standard input. 
Every Dodona Docker image shares a `main.sh` file that passes through this information to the actual testing framework, while setting up some error handling.
The testing process in the Docker containers sends back the test results by writing a JSON document to its standard output channel.
In June 2019, R support was added to Dodona using an image derived from the `rocker/r-base` image that sets up the `runner` user and and `main.sh` file expected by Dodona\footnote{\href{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}}. 
It also installs the packages required for the testing framework and the exercises so that this does not have to happen every time a student's submission is evaluated.
The actual testing of R exercises is done using a custom framework loosely based on \CRANpkg{testthat} \citep{wickham_testthat_2011}.
During the development of the testing framework, it was found that the \CRANpkg{testthat} framework did not provide enough information to its reporter system to send back all the fields required by Dodona to render its feedback.
Right now, multiple statistics courses are developing exercises to automate the feedback for their lab classes.
